---
title: "Iteratively reweighted least squares"
subtitle: "BMLR Chapter 5"
format: 
  revealjs:
    output-file: "05_iwls_ch5.html"
    slide-number: true
  html:
    output-file: "05_iwls_ch5_o.html"
logo: "../img/favicon.png"
---

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(tidymodels)
library(GGally)
library(knitr)
library(patchwork)
library(viridis)
library(ggfortify)
library(gridExtra)
```



# Iteratively reweighted least squares (IWLS)

## Data: Noisy Miners {.smaller}

The dataset [`nminer`](data/nminer.csv) contains information about the number of noisy miners (small Australian bird) detected in two woodland patches within the Wimmera Plains of Victoria, Australia. It was obtained from the **GLMsdata** R package. We will use the following variables:

 
- **`Minerab`**: The number of noisy miners (abundance) observed in three 20 minute surveys
- **`Eucs`**: The number of eucalyptus trees in each 2 hectare area (about 4.94 acres)




## Noisy Miner Model {.smaller}


```{r echo = F}
nminer <- read_csv("data/nminer.csv")
nminer %>%
  slice(1:5) %>% select(Eucs, Minerab) %>% kable(digits = 1)
```

. . .

Our goal is to use a Poisson regression model to predict the number of noisy miners observed in three 20 minute surveys based on the number of eucalyptus trees.

$$\log(\lambda_{Minearab}) = \beta_0 + \beta_1 ~ Euc$$

. . .

**What are the best estimates of $\beta_0$ and $\beta_1$?**




## Iteratively reweighted least squares (IWLS) {.smaller}

- The estimates of $\beta_0$ and $\beta_1$ are found using maximum likelihood estimation. 


- **Iteratively reweighted least-squares (IWLS)** is used to find the MLEs
  - Nelder and Wedderburn (1972) show that under certain specifications of the weights and a modified response variable, the estimates found using IWLS are equivalent to the MLEs.
  


## IWLS Set up {.smaller}

**Working response**: Modified response variable at each step of the iteration. 

$$z_i = g(\theta) + g'(\theta)(y_i - \theta_i)$$

For Poisson regression, this is 

$$z_i = \log(\lambda) + \frac{(y_i - \lambda_i)}{\lambda_i}$$

. . .

**Working Weights:** Weights applied to the observations at each step of the iteration 

$$W_i = \frac{\theta^2}{Var(Y)} \hspace{5mm} \Rightarrow \hspace{5mm}  W_i = \frac{\lambda^2}{\lambda} =  \lambda \text{ for Poisson regression}$$



## IWLS procedure {.smaller}

1. Find initial starting values $\hat{\theta}_i$.

2. Calculate the working response values $z_i$.

3. Calculate the working weights $W_i$.

4. Find the coefficient estimates of the weighted least squares model.

. . .

$z_i = \beta_0 + \beta_1 x \hspace{5mm} \text{ with weights }W_i$

The estimates $\hat{\beta}_0$ and $\hat{\beta}_1$ are the estimates for the model coefficients. 

. . .

Use $\hat{\beta}_0$ and $\hat{\beta}_1$ to calculate updated values of $\hat{\theta}_i$ and repeat steps 2 - 4 until convergence.



. . .

## Demo in `ch5_iwls.R` in server class files


## Acknowledgements {.smaller}

- These slides are based on content in [BMLR: Chapter 4](https://bookdown.org/roback/bookdown-BeyondMLR/)

- Initial versions of the slides are by Dr. Maria Tackett, Duke University

- [BMLR: Chapter 5 - Generalized Linear Models: A Unifying Theory](https://bookdown.org/roback/bookdown-BeyondMLR/ch-glms.html)
- Nelder, J. A., & Wedderburn, R. W. (1972). Generalized linear models. Journal of the Royal Statistical Society: Series A (General), 135(3), 370-384.
- Generalized Linear Models with Examples in R
  - Chapter 5 - Generalized Linear Models: Structure
  - Chapter 6 - Generalized Linear Models: Estimation 
  