---
title: Homework 2
subtitle: 'Chapter 7: Moving Beyond Linearity, ISLR'
---

`r if (knitr:::is_html_output()) '\\newcommand{\\E}{\\operatorname{E}}'`
`r if (knitr:::is_html_output()) '\\newcommand{\\SD}{\\operatorname{SD}}'`
`r if (knitr:::is_html_output()) '\\newcommand{\\var}{\\operatorname{Var}}'`


## Setup

Each of your assignments will begin with the following steps.

-   Going to our RStudio Server at <https://csgateway.cornellcollege.edu:8787/>

-   Creating a new *R project*, inside your homework folder on the server, and giving it a sensible name such as *homework_2* and having that project in the course folder you created. 

-   Create a new quarto document and give it a sensible name such as *hw2*.

-   In the `YAML` add the following (add what you don't have). The embed-resources component will make your final rendered `html` self-contained.

```{r}
#| label: yaml_example
#| eval: false
---
title: "Document title"
author: "my name"
format:
  html:
    embed-resources: true
---
```

## Instructions

Be sure to include the relevant R code as well as full sentences answering each of the questions (i.e. if I ask for the average, you can output the answer in R but **also** write a full sentence with the answer). Be sure to frequently save your files!

In this homework we will work with four packages: `tidyverse` which is a collection of packages for doing data analysis in a "tidy" way, `tidymodels` for statistical model coefficients, `splines` for our splines models and `ISLR2` for the `Boston` data. 


## Code

#### Polynomial Regression

To fit a degree 4 polynomial regression model you use:

```{r}
#| eval: false
model1 <- lm(wage ~ poly(age , 4, raw = TRUE), data = wage)
```

#### Cubic Spline

To fit a cubic spline (degree 3) polynomial regression model you use:

```{r}
#| eval: false
model2<- lm(wage ~ bs(age , knots = c(22, 33, 50), degree = 3), data = Wage)
```

Use quantiles of the predictor to choose your knots. The choice should be motivated by some EDA. 


#### Natual Cubic Spline

To fit a degree natural cubic spline. 

```{r}
#| eval: false
model3 <- lm(wage ~ ns(age,df = 4), data = Wage)
```

Instead of providing the degrees of freedom (df), you can provide knots. 


## Exercises

All problems are from {{< var course.text2 >}} Abbreviated ISLR. 

```{r}
library(ISLR2)
library(ggplot2)
library(splines)
library(tidymodels)
```


1. (ISLR, Chapter 7, Exercise 3) Suppose we fit a curve with basis functions $b_1(X) = X,\space b_2(X) = (X − 1)^2I(X \geq 1)$. (Note that $I(X \geq 1)$ equals 1 for $X \geq 1$ and 0 otherwise.) We fit the linear regression model 

$$Y = \beta_0 + \beta_1b_1(X) + \beta_2b_2(X) + \epsilon,$$ <br>

and obtain coefficient estimates $\hat{\beta}_0 = 1,\space \hat{\beta}_1 = 1,\space \hat{\beta}_2 = −2$. Sketch the estimated curve between $X = −2$ and $X = 2$. Note the intercepts, slopes, and other relevant information.

2. (ISLR, Chapter 7, Exercise 9, edited) This question uses the variables `dis` (the weighted mean of distances to five Boston employment centers) and `nox` (nitrogen oxides concentration in parts per 10 million) from the Boston data (in package `ISLR2`). We will treat `dis` as the predictor and `nox` as the response.

  (a) Use the `poly()` function to fit a cubic polynomial regression to predict `nox` using `dis`. Report the regression output, and plot the resulting data and polynomial fits.
  
```{r}

model1 <- lm(nox ~ poly(dis , 3, raw = TRUE), data = Boston)
summary(model1)
anova(model1)
ggplot(Boston, aes(x=dis, y=nox))+
  geom_point()+
  geom_smooth(formula=y~poly(x,4), method="lm")
  

```
  
  (b) Plot the polynomial fits for a range of different polynomial degrees (say, from 1 to 5), and report the associated residual sum of squares.
  
```{r}
ggplot(Boston, aes(x=dis, y=nox))+
  geom_point()+
  geom_smooth(formula=y~poly(x,1), method="lm")
ggplot(Boston, aes(x=dis, y=nox))+
  geom_point()+
  geom_smooth(formula=y~poly(x,2), method="lm")
ggplot(Boston, aes(x=dis, y=nox))+
  geom_point()+
  geom_smooth(formula=y~poly(x,3), method="lm")
ggplot(Boston, aes(x=dis, y=nox))+
  geom_point()+
  geom_smooth(formula=y~poly(x,4), method="lm")
ggplot(Boston, aes(x=dis, y=nox))+
  geom_point()+
  geom_smooth(formula=y~poly(x,5), method="lm")
anova(lm(nox ~ poly(dis , 1, raw = TRUE), data = Boston))
anova(lm(nox ~ poly(dis , 2, raw = TRUE), data = Boston))
anova(lm(nox ~ poly(dis , 3, raw = TRUE), data = Boston))
anova(lm(nox ~ poly(dis , 4, raw = TRUE), data = Boston))
anova(lm(nox ~ poly(dis , 5, raw = TRUE), data = Boston))
```
  
  (c) Select the optimal degree for the polynomial, and explain your results.
  
  Third degree polynomial has a good fit and increasing to 4 or 5 does not appreciably reduces the residual sum of squares.
  
  (d) Use the `bs()` function to fit a regression spline to predict `nox` using `dis`. Report the output for the fit using four degrees of freedom. How did you choose the knots? Plot the resulting fit.
  
  I let the computer pick the knots. They are based on quantiles.
  
```{r}
model2<- lm(nox ~ bs(dis, df=4), data = Boston)
summary(model2)

ggplot(data=Boston, aes(x=dis, y=nox)) +
  geom_point(aes(x=dis, y=nox)) +
  geom_smooth(formula =y ~ bs(x, df=4), method="lm")


```
  
  (e) Now fit a regression spline for a range of degrees of freedom, and plot the resulting fits and report the resulting RSS. Describe the results obtained.
  
```{r}
anova(lm(nox ~ bs(dis, df=2), data = Boston))

ggplot(data=Boston, aes(x=dis, y=nox)) +
  geom_point(aes(x=dis, y=nox)) +
  geom_smooth(formula =y ~ bs(x, df=2), method="lm")

anova(lm(nox ~ bs(dis, df=3), data = Boston))

ggplot(data=Boston, aes(x=dis, y=nox)) +
  geom_point(aes(x=dis, y=nox)) +
  geom_smooth(formula =y ~ bs(x, df=3), method="lm")

anova(lm(nox ~ bs(dis, df=4), data = Boston))

ggplot(data=Boston, aes(x=dis, y=nox)) +
  geom_point(aes(x=dis, y=nox)) +
  geom_smooth(formula =y ~ bs(x, df=4), method="lm")


anova(lm(nox ~ bs(dis, df=5), data = Boston))

ggplot(data=Boston, aes(x=dis, y=nox)) +
  geom_point(aes(x=dis, y=nox)) +
  geom_smooth(formula =y ~ bs(x, df=5), method="lm")

anova(lm(nox ~ bs(dis, df=6), data = Boston))

ggplot(data=Boston, aes(x=dis, y=nox)) +
  geom_point(aes(x=dis, y=nox)) +
  geom_smooth(formula =y ~ bs(x, df=6), method="lm")

anova(lm(nox ~ bs(dis, df=7), data = Boston))

ggplot(data=Boston, aes(x=dis, y=nox)) +
  geom_point(aes(x=dis, y=nox)) +
  geom_smooth(formula =y ~ bs(x, df=7), method="lm")





```  

Choose the model with df=5 for the biggest drop in RSS.  After that changes in RSS are very small.

  (e) Now fit a natural regression spline for a range of degrees of freedom, and plot the resulting fits and report the resulting RSS. Describe the results obtained.
  
```{r}

anova(lm(nox ~ ns(dis, df=3), data = Boston))

ggplot(data=Boston, aes(x=dis, y=nox)) +
  geom_point(aes(x=dis, y=nox)) +
  geom_smooth(formula =y ~ ns(x, df=3), method="lm")

anova(lm(nox ~ ns(dis, df=4), data = Boston))

ggplot(data=Boston, aes(x=dis, y=nox)) +
  geom_point(aes(x=dis, y=nox)) +
  geom_smooth(formula =y ~ ns(x, df=4), method="lm")


anova(lm(nox ~ ns(dis, df=5), data = Boston))

ggplot(data=Boston, aes(x=dis, y=nox)) +
  geom_point(aes(x=dis, y=nox)) +
  geom_smooth(formula =y ~ ns(x, df=5), method="lm")

anova(lm(nox ~ ns(dis, df=6), data = Boston))

ggplot(data=Boston, aes(x=dis, y=nox)) +
  geom_point(aes(x=dis, y=nox)) +
  geom_smooth(formula =y ~ ns(x, df=6), method="lm")

anova(lm(nox ~ ns(dis, df=7), data = Boston))

ggplot(data=Boston, aes(x=dis, y=nox)) +
  geom_point(aes(x=dis, y=nox)) +
  geom_smooth(formula =y ~ ns(x, df=7), method="lm")





```  
  
  (f) Pick an overall best model using various metrics, complexity, and regression assumptions to make your choice. They are the same metrics used in multiple linear regression. 
  
```{r}
AIC(lm(nox ~ bs(dis, df=3), data = Boston))
AIC(lm(nox ~ bs(dis, df=4), data = Boston))
AIC(lm(nox ~ bs(dis, df=5), data = Boston))
AIC(lm(nox ~ bs(dis, df=6), data = Boston))
AIC(lm(nox ~ bs(dis, df=7), data = Boston))
AIC(lm(nox ~ ns(dis, df=3), data = Boston))
AIC(lm(nox ~ ns(dis, df=4), data = Boston))
AIC(lm(nox ~ ns(dis, df=5), data = Boston))
AIC(lm(nox ~ ns(dis, df=6), data = Boston))
AIC(lm(nox ~ ns(dis, df=7), data = Boston))

BIC(lm(nox ~ bs(dis, df=3), data = Boston))
BIC(lm(nox ~ bs(dis, df=4), data = Boston))
BIC(lm(nox ~ bs(dis, df=5), data = Boston))
BIC(lm(nox ~ bs(dis, df=6), data = Boston))
BIC(lm(nox ~ bs(dis, df=7), data = Boston))
BIC(lm(nox ~ ns(dis, df=3), data = Boston))
BIC(lm(nox ~ ns(dis, df=4), data = Boston))
BIC(lm(nox ~ ns(dis, df=5), data = Boston))
BIC(lm(nox ~ ns(dis, df=6), data = Boston))
BIC(lm(nox ~ ns(dis, df=7), data = Boston))

summary(lm(nox ~ bs(dis, df=3), data = Boston))
summary(lm(nox ~ bs(dis, df=4), data = Boston))
summary(lm(nox ~ bs(dis, df=5), data = Boston))
summary(lm(nox ~ bs(dis, df=6), data = Boston))
summary(lm(nox ~ bs(dis, df=7), data = Boston))
summary(lm(nox ~ ns(dis, df=3), data = Boston))
summary(lm(nox ~ ns(dis, df=4), data = Boston))
summary(lm(nox ~ ns(dis, df=5), data = Boston))
summary(lm(nox ~ ns(dis, df=6), data = Boston))
summary(lm(nox ~ ns(dis, df=7), data = Boston))

```

NS 4 looks like the best (do degree 5)

# Submission

When you are finished with your homework, be sure to Render the final document. Once rendered, you can download your file by:

-   Finding the .html file in your File pane (on the bottom right of the screen)
-   Click the check box next to the file
-   Click the blue gear above and then click "Export" to download
-   Submit your final html document to the respective assignment on [Moodle]({{< var course.moodle >}})



