---
title: "Lab 02 SeaIce and Splines"
format: html
---


## Lab Setup

1) Copy the project lab folder located at *Home -> STA363_inst_files-> labs*. If you check the box next to the folder name, then click the small gear icon you can "copy to" and put a copy of the folder in your newly created folder. 
2) Now, click *Project:(None)* in the top right corner of your screen and choose *New Project* and "Don't Save". 
3) Choose *Existing Directory* and navigate to the directory you just copied.
4) You can place your responses in the *qmd* file included. (There is only one *qmd* file in this directory.)

## Packages

In this lab, we will work with three packages: `tidyverse`, which is a collection of packages for data analysis in a "tidy" way; `tidymodels` for statistical model coefficients; and `splines` for our spline models.

If you'd like to run your code in the Console as well, you'll need to load the packages there as well. To do so, run the following in the console.

```{r output=FALSE}
library(tidyverse) 
library(tidymodels)
library(splines)
```

## Data

For this lab, we are using `seaice` data downloaded from an NOAA webiste. This can be loaded in by running:

```{r output=FALSE}
seaice <- read_csv("data/seaice.csv")
```

Below is a data dictionary for this dataset:

| variable   | definition                                                                                                                      |
|----------------|-----------------------------------------------------------|
| year           | Year that ice information was recorded                                                                                         |
| mo             | Month data was recorded - September in all cases here                                                                           |
| source_dataset | Which NOAA dataset the data came from (all but the most recent observation come from the same dataset)                     |
| region         | All observations come from the N(orth) or artic region                                                                      |
| extent         | Total region with at least 15% ice cover                                                                                    |
| area           | Total region covered by ice                                                                                                 |

Before we run any regression models, we should center the year variable (as we did in class).  Here we create the new variable `c_year`.

```{r}
seaice <- seaice |>
  mutate(c_year = (year-1979))
```

### Polynomial Regression

Polynomial regression can be thought of as doing polynomial expansion on a variable and passing that expansion into a linear regression model. We will be very explicit in this formulation in this chapter.

The following step will take `c_year` and replace it with the variables `c_year`, `c_year^2`, `c_year^3`, and `c_year^4` since we set `degree = 4` (actually, that's not quite true, keep reading).

```{r}
model1 <- lm(extent ~ poly(c_year , 4), data = seaice)
coef(summary(model1))
```

Though we expect `age`, `age^2`, `age^3`, and `age^4`, the function it returns uses variables that are a "basis of orthogonal polynomials".  That is a mathematical way of saying that each of the columns is a linear combination of the variables `age`, `age^2`, `age^3`, and `age^4`. We can see this by using `poly()` directly with `raw = FALSE` since it is the default

```{r}
poly(1:6, degree = 4, raw = FALSE)
```

These variables don't have the format we assumed. But this is still a well-reasoned (mathematically sound) transformation (see MAT 221). We can get the raw polynomial transformation by setting `raw = TRUE`

```{r}
poly(1:6, degree = 4, raw = TRUE)
```

These transformations align with what we would expect. It is still recommended to stick with the default of `raw = FALSE` unless you have a reason not to do that.  Just realize that the interpretation will not be as straight-forward.  But then the interpretation of a polynomial is not really straight-forward anyway.

# Polynomial Regression Exercises

In class we explored the `extent` variable.  For this lab, you will explore the `area` variable. 

1.  For this lab, we are interested in predicting the variable `area` from `c_year`. Complete a brief bivariate EDA for `c_year` and the response.

2.  Using your EDA from the last step, what degree polynomial do you think would be appropriate `c_year` vs `area`?

5.  Fit a polynomial regression model with your chosen degree of the polynomial for `area` to predict `c_year`.

6.  We need to choose the "best" degree polynomial regression model, and one way is hypothesis testing. Fit polynomial regression models with `year` predicting `area` with degrees 1 through 5.

7.  Use either a drop-in deviance test or a Nested F Test to determine which model is best.

8.  Now create a plot that visualizes the fit of the data. Make sure to include the original data values and a line for the fit polynomial model.

### Splines

To fit regression splines, we use the splines library. We saw that regression splines can be fit by constructing an appropriate matrix of basis functions. The `bs()` function generates the entire matrix of basis functions for splines with the specified set of knots. By default, cubic `bs()` splines are produced. Fitting `area` to `c_year` using a regression spline is simple:

```{r}
#| eval: false
model-spline<- lm(area ~ bs(c_year , knots = c(a, b, c)), data = seaice)
```

Here we have pre-specified knots at ages $a$, $b$, and $c$. This will produce a spline with six basis functions. (Recall that a cubic spline with three knots has seven degrees of freedom; these degrees of freedom are used up by an intercept, plus six basis functions.) We could also use the df option to produce a spline with knots at uniform quantiles of the data.

### Splines Exercises 

9.  Use your EDA or new data visualizations to choose the 3 knots. Then fit the cubic spline model using your chosen knots.

10. Another way to find the knots is quantiles. Use the `bs` function, specifying the correct number of degrees of freedom, to fit the cubic spline model using knots at the 25th, 50th, and 75th quantiles.

11. To fit a natural spline, we use the `ns()` function. Here we fit a natural spline with four degrees of freedom. Change the code below to fit a natural cubic spline.

```{r}
#| eval: false
model3 <- lm(area ~ ns(c_year , df = d), data = seaice)
```


13. Now we want to choose the best model from all the models above. Use AIC, BIC, and RMSE (the square root of mean square error), model complexity, and residual plots, to choose the best model. Explicitly explain your reasoning.


